# Reproducible Research: Peer Assessment 1
Brian Lewis

## Loading and preprocessing the data

```{r loaddata}
echo = TRUE
ActMon <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

```{r centraltendancy}
echo = TRUE
total.steps <- tapply(ActMon$steps, ActMon$date, sum, na.rm = TRUE)
mean <- mean(total.steps)
print(mean)
median <- median(total.steps)
print(median)
```

### Draw Histogram

```{r histogram}
echo = TRUE
library(ggplot2)
qplot(total.steps, xlab = "Total Steps", ylab = "Frequency", main = "Histogram of Daily Steps", fill = "vermillion")
```

## What is the average daily activity pattern?

### Create date.time field and repopulate time field to analyze time intervals

```{r time field manipulation}
echo = TRUE
time <- formatC(ActMon$interval/100, 2, format = "f")
ActMon$date.time <- as.POSIXct(paste(ActMon$date, time), format = "%Y-%m-%d %H.%M", 
    tz = "GMT")
ActMon$time <- format(ActMon$date.time, format = "%H:%M:%S")
ActMon$time <- as.POSIXct(ActMon$time, format = "%H:%M:%S")
```


### Calculate the mean steps for each five minute interval.

```{r mean steps interval}
echo = TRUE
mean.steps <- tapply(ActMon$steps, ActMon$time, mean, na.rm = TRUE)
daily.pattern <- data.frame(time = as.POSIXct(names(mean.steps)), mean.steps = mean.steps)
```

### Produce time series plot for the mean steps.

```{r mean steps time plot}
echo = TRUE
library(scales)
ggplot(daily.pattern, aes(time, mean.steps)) + geom_line() + xlab("Time of Day") + 
    ylab("Mean Number of Steps") + ggtitle("Time Series Plot for Mean Steps") + scale_x_datetime(labels = date_format(format = "%H:%M"))
```

### Which five minute interval has the highest mean number of steps?

```{r interval with highest mean}
echo = TRUE
highest <- which.max(daily.pattern$mean.steps)
format(daily.pattern[highest, "time"], format = "%H:%M")
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r number of NAs}
echo = TRUE
sum(is.na(ActMon$steps))
```

### Impute missing steps using the 5 minute interval mean.

```{r impute NAs}
echo = TRUE
library(Hmisc)
ActMon.imputed <- ActMon
ActMon.imputed$steps <- with(ActMon.imputed, impute(steps, mean))
```

### Compare means and medians of original and imputed datasets.

```{r imputed mean and median}
echo = TRUE
total.steps.imputed <- tapply(ActMon.imputed$steps, ActMon.imputed$date, 
    sum)
mean(total.steps)
mean(total.steps.imputed)
median(total.steps)
median(total.steps.imputed)
```

### Both the central tendancy measures, mean and median, increased when the NA values were imputed.

### Draw histogram of imputed dataset.

```{r imputed histogram}
echo = TRUE
library(ggplot2)
qplot(total.steps.imputed, xlab = "Total Steps", ylab = "Frequency", main = "Histogram of Daily Steps", fill = "darkorchid4")
```

## Are there differences in activity patterns between weekdays and weekends?

### Add a field indicating whether a day falls on a weekday or weekend.

```{r}
day.type <- function(date) {
    if (weekdays(date) %in% c("Saturday", "Sunday")) {
        return("weekend")
    } else {
        return("weekday")
    }
}
day.types <- sapply(ActMon.imputed$date.time, day.type)
ActMon.imputed$day.type <- as.factor(day.types)
```

### Create a dataset that contains the weekday and weekend mean steps data.

```{r}
mean.steps <- tapply(ActMon.imputed$steps, interaction(ActMon.imputed$time, 
    ActMon.imputed$day.type), mean, na.rm = TRUE)
day.type.pattern <- data.frame(time = as.POSIXct(names(mean.steps)), mean.steps = mean.steps, 
    day.type = as.factor(c(rep("weekday", 288), rep("weekend", 288))))
```

### Draw histograms comparing weekday and weekend means.

```{r}
ggplot(day.type.pattern, aes(time, mean.steps)) + geom_line() + xlab("Time of day") + 
    ylab("Mean number of steps") + scale_x_datetime(labels = date_format(format = "%H:%M")) + 
    facet_grid(. ~ day.type)
```
